Poker Chips, Earthquakes, and Computation

### 2014-09-27 draft0.tex
Edited Section "Time Until Stability"

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{verbatim}
\usepackage[margin=0.9in]{geometry}
%\usepackage{amsfonts}
%\usepackage{amsthm}
\usepackage{subfigure}
%\usepackage{caption}
%\usepackage{subcaption}
%\usepackage{cite}
\usepackage{hyperref}
\usepackage{url}
\urlstyle{same}
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

@@

We want to be able to bound the number of global earthquakes that it takes for a finite board to reach a stable state. We will first focus on the case of the board with only one stack containing $n$ chips. Without loss of generality, lets assume that the board contains $n$ chips at location $(0,0)$. We define $T: \mathbb{Z} \rightarrow \mathbb{Z}$ where $T(n)$ denotes the number of global earthquakes needed until the board with one stack of $n$ chips stabilizes. We also let $T':\mathbb{Z} \rightarrow \mathbb{Z}$ where $T'(n)$ denotes the local earthquakes that must hit $(0,0)$ in order for the board to stabilize. Its clear that $T(n) \geq T'(n)$. We also know that $T'(n) \geq \lfloor \frac{n}{4}\rfloor$.

One can get another trivial upper bound for $T(n)$ by looking at the maximum value of the potential function. Remember that the board will only contain chips in the areas denotes by locations in $[-\sqrt{n}, \sqrt{n}] \times [-\sqrt{n}, \sqrt{n}]$. So we know that when the board is stable at some $f'$, 
\[ \Phi(f') \leq \sum_{x,y} f'(x,y)(|x| + |y|)^2 \leq 4n*3*4n = O(n^2) \]
And since $\Phi(f) = 0$ and we know that $\Phi$ increases by at least $4$ in each global earthquake, 
\[ T(n) = O(n^2) \]
We can get tighter bounds by looking at the number of earthquakes more carefully. Lets let $T(x,y)$ be the number of local earthquakes needed to hit location $(x,y)$ for the board to stabilize. Suppose $(x,y) \neq (0,0)$, then the number of earthquakes required to hit location $(x,y)$ will be the total number of chips entering location $(x,y)$ divided by 4 (taking floors ofcourse). In particular,
\[ T(x,y) = \left\{ \begin{array}{cc} \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor & (x,y) \neq (0,0) \\
 \lfloor \frac{n}{4} \rfloor  + \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor & (x,y) = (0,0) \end{array} \right. \]
So solving for $T$ is a solution to a discrete version of the Laplacian. We were able to experiment different scenarious to understand the growth of $T(n)$. 
\begin{figure}[!ht]
\centering
\subfigure[Global earthquakes until stable. Horizontal axis is $n$ and vertical axis is the number of global earthquakes]{\includegraphics[width=0.4\linewidth]{TimeUntilStabilizeWithOnePeak}\label{fig:case1}} \qquad
\subfigure[Global earthquakes until stable. Horizontal axis is $n$ and vertical axis is the number of global earthquakes divided by $n$]{\includegraphics[width=0.4\linewidth]{Timeuntilstabledividedbyn}\label{fig:case2}}
\caption{Number of global earthquakes until stability}
\label{fig:growthofT}
\end{figure}
By looking at Figure~\ref{fig:growthofT} seems that $T(n) = \Theta(n\log n)$. Our strategy to show a lower bound on $T(n)$ by giving a lower bound on $T'(n)$ seems to be the correct approach, since by experimental evidence (see Figure~\ref{fig:growthofT'}, we believe that $T'(n) = \Theta(n\log n)$.

>>>>

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{verbatim}
\usepackage[margin=0.9in]{geometry}
%\usepackage{amsfonts}
%\usepackage{amsthm}
\usepackage{subfigure}
\usepackage{mathtools}
%\usepackage{caption}
%\usepackage{subcaption}
%\usepackage{cite}
\usepackage{hyperref}
\usepackage{url}
\urlstyle{same}
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

@@

Another question of interest is how many global earthquakes it would take for a finite board to reach a stable state. We will first focus on the case of the board with only one stack containing $n$ chips.

Without loss of generality, we can assume that the board contains $n$ chips at location $(0,0)$. We first make two definitions we will use in our discussion.
\begin{definition}
We define $T: \mathbb{Z} \rightarrow \mathbb{Z}$ where $T(n)$ denotes the number of global earthquakes needed until the board with one stack of $n$ chips stabilizes. We also define $T':\mathbb{Z} \rightarrow \mathbb{Z}$ where $T'(n)$ denotes the local earthquakes that must hit $(0,0)$ in order for the board to stabilize. 
\end{definition}
Its clear that $T(n) \geq T'(n)$. We also know that $T'(n) \geq \lfloor \frac{n}{4}\rfloor$, since that many earthquakes will reduce the number of chips on $(0,0)$ to less than $4$.

We can get another upper bound for $T(n)$ by looking at the maximum value of the potential function. Recall that the board will be nonzero only in locations in the range $[-\sqrt{n}, \sqrt{n}] \times [-\sqrt{n}, \sqrt{n}]$. 
So we know that when the board is stable at some $f'$, 
\begin{equation*}
\Phi(f') \leq \sum_{x,y} f'(x,y)(|x| + |y|)^2 \leq \sum_{x,y} 3*4n = 3*4n*4n= O(n^2)
\end{equation*}
We know that $\Phi(f) = 0$ and that $\Phi$ increases by at least $4$ after each global earthquake, so
\[ T(n) = O(n^2) \]
We can get tighter bounds by looking at the number of earthquakes more carefully.
\begin{definition}
 Let $T(x,y)$ be the number of local earthquakes needed to hit location $(x,y)$ for the board to stabilize. 
 \end{definition}
 If $(x,y) \neq (0,0)$, then the number of earthquakes required to hit location $(x,y)$ will be the total number of chips entering location $(x,y)$ divided by 4 (after taking floors). For $(0,0)$ the number of earthquakes that must hit the square equals the number of chips entering $(x,y)$ divided by 4 (after taking floors) plus the number of earthquakes needed to clear the original $n$ chips off the square.
 Since a chip only enters a square when an earthquake hits an adjacent square, we have the following formula
\begin{equation*}
T(x,y) = \begin{dcases} \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  (x,y) \neq (0,0) \\
 \lfloor \frac{n}{4} \rfloor  + \lfloor\frac{T(x+1, y)}{4} \rfloor + \lfloor \frac{T(x-1, y)}{4} \rfloor + \lfloor \frac{T(x, y-1)}{4} \rfloor + \lfloor \frac{T(x, y+1)}{4} \rfloor  (x,y) = (0,0) .
 \end{dcases}
 \end{equation*}
So solving for $T$ is a solution to a discrete version of the Laplacian. We were able to experiment different scenarious to understand the growth of $T(n)$. 
\begin{figure}[!ht]
\centering
\subfigure[Global earthquakes until stable. Horizontal axis is $n$ and vertical axis is the number of global earthquakes]{\includegraphics[width=0.4\linewidth]{TimeUntilStabilizeWithOnePeak}\label{fig:case1}} \qquad
\subfigure[Global earthquakes until stable. Horizontal axis is $n$ and vertical axis is the number of global earthquakes divided by $n$]{\includegraphics[width=0.4\linewidth]{Timeuntilstabledividedbyn}\label{fig:case2}}
\caption{Number of global earthquakes until stability}
\label{fig:growthofT}
\end{figure}
By looking at Figure~\ref{fig:growthofT} it seems that $T(n) = \Theta(n\log n)$. Our strategy to show a lower bound on $T(n)$ by giving a lower bound on $T'(n)$ seems to be the correct approach, since by experimental evidence (see Figure~\ref{fig:growthofT'}, we believe that $T'(n) = \Theta(n\log n)$.
